WASP-4 is Accelerating Toward the Earth

### 2020-02-23 paper/ms.tex
minor
text update

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% TODO: kishore reference
% TODO: rachel comments
% TODO: steve comments

@@

This study draws attention to a confounding factor that, while
elementary, does not seem to have received the attention it deserves.
The point is that observational programs aimed at identifying orbital
decay in hot Jupiters through transit timing must be accompanied by
concurrent long-term radial velocity monitoring.  The reason is that
an apparent change in period can be produced by the Doppler effect
associated with acceleration of the hot-Jupiter host star along the
line of sight, such as the acceleration that might be produced by a
massive, wide-orbiting companion.  Massive outer companions to hot
Jupiters are common.  \citet{bryan_statistics_2016} calculated an
occurrence rate of $60.9^{+5.2}_{-5.6}\%$ for outer companions to hot
Jupiters with masses from 1-20$\,$$M_{\rm Jup}$ and semi-major axes
from 5-100$\,$AU.  Therefore, we expect that many hot Jupiters will
display secular trends in orbital period that are unrelated to tidal
orbital decay.  This possibility can be checked by performing
long-term radial-velocity monitoring at a level sensitive enough to
detect or rule out the relevant amplitude of acceleration.

@@

Based on the available data, the companion causing the acceleration is
probably either a brown-dwarf or low mass star with an orbital
distance of $10$-$100\,{\rm AU}$ (Figure~\ref{fig:mass_sma}).  Given
such a mass, this companion may have influenced the orbital evolution
of the hot Jupiter orbiting WASP-4, as well as any other planets.  The
fact that most hot Jupiters have similar massive outer companions
\citep{knutson_friends_2014,bryan_statistics_2016} is compatible with
some of the high-eccentricity formation theories for hot Jupiters (see
\citealt{dawson_johnson_2018}).  Further radial velocity monitoring
should eventually reveal the orbital parameters and minimum mass of
WASP-4's companion.

@@

\begin{figure}[t]
	\begin{center}
		\leavevmode
		\includegraphics[width=0.48\textwidth]{f5.pdf}
	\end{center}
	\vspace{-0.4cm}
	\caption{
  {\bf Predicted hot Jupiter period changes from linear radial
  velocity trends.} Including WASP-4b, 16 of 51 hot Jupiters from
  \citet{knutson_friends_2014} have shown long-term radial velocity
  trends.  HAT-P-11 is shown, though its signal may be caused by
  stellar activity.  Three hot Jupiters are not shown because their
  radial velocity curves are better described as quadratic trends in
  time: HAT-P-17, WASP-8, and WASP-34.  Objects are ordered in the $y$
  dimension by the absolute value of d$P$/d$t$.
	\label{fig:pdot_pop}
  \vspace{-0.3cm}
	}
\end{figure}

>>>>

@@

This study draws attention to a confounding factor that, while
elementary, does not seem to have received the attention it deserves.
The point is that observational programs aimed at identifying orbital
decay in hot Jupiters through transit timing must be accompanied by
concurrent long-term radial velocity monitoring.  The reason is that
an apparent change in period can be produced by the Doppler effect
associated with acceleration of the hot-Jupiter host star along the
line of sight, such as the acceleration that might be produced by a
massive, wide-orbiting companion.  Massive outer companions to hot
Jupiters are common.  \citet{bryan_statistics_2016} calculated an
occurrence rate of $70\pm8\%$ for outer companions to hot Jupiters
with masses from 1-13$\,$$M_{\rm Jup}$ and semi-major axes from
1-20$\,$AU.  Therefore, we expect that many hot Jupiters will display
secular trends in orbital period that are unrelated to tidal orbital
decay.  This possibility can be checked by performing long-term
radial-velocity monitoring at a level sensitive enough to detect or
rule out the relevant amplitude of acceleration.

@@

Based on the data, the companion causing the acceleration is probably
either a brown-dwarf or low mass star with an orbital distance of
$10$-$100\,{\rm AU}$ (Figure~\ref{fig:mass_sma}).  Given such a mass,
this companion may have influenced the orbital evolution of the hot
Jupiter orbiting WASP-4, as well as any other planets in the system.
The fact that most hot Jupiters have similar massive outer companions
\citep{knutson_friends_2014,bryan_statistics_2016} is compatible with
some of the high-eccentricity formation theories for hot Jupiters (see
\citealt{dawson_johnson_2018}).  Further radial velocity monitoring
should eventually reveal the orbital parameters and minimum mass of
WASP-4's massive outer companion.

Based on a priori expectations, if the outer companion lives within
100$\,$AU, then it seems more likely to be a brown-dwarf than a
low-mass star.  The reason is that stellar companions within 100$\,$AU
seem to be rare in systems with transiting planets, relative to field
stars.  This issue has been reviewed by \citet{moe_impact_2019}, who
synthesized work by \citet{wang_influence_2014,wang_influence_2015},
\citet{kraus_impact_2016}, \citet{ngo_2015_fohj2}, \citet{ngo_friends_2016},
\citet{matson_stellar_2018}, \citet{ziegler_soar_2020}, and others.
A handful of systems with inner hot Jupiters and outer brown dwarfs within
100$\,$AU are known, for instance CoRoT-20, HATS-59, WASP-53,
and WASP-81
\citep{triaud_peculiar_2017,rey_brown_2018,sarkis_hats-59bc_2018}.  In
contrast, from surveys by
\citet{knutson_friends_2014}, \citet{ngo_2015_fohj2} and
\citet{mugrauer_search_2019} we could find only one example of a
system with an inner hot Jupiter and an outer low-mass star within
100$\,$AU: HAT-P-10 \citep{bakos_2009_hatp10}.  Overall, this body of
literature indicates that the presence of a stellar-mass companion
within 100$\,$AU of the host star could hinder the formation of
planetary systems.  If true, then one would expect that the outer
companion in WASP-4 would have the lowest mass allowed by the data.

@@

\begin{figure}[t]
	\begin{center}
		\leavevmode
		\includegraphics[width=0.48\textwidth]{f5.pdf}
	\end{center}
	\vspace{-0.4cm}
	\caption{
  {\bf Predicted hot Jupiter period changes from linear radial
  velocity trends.} Including WASP-4b, 16 of 51 hot Jupiters from
  \citet{knutson_friends_2014} have shown long-term radial velocity
  trends.  HAT-P-11 is shown, though its signal is somewhat correlated
  with stellar activity \citep{yee_hat-p-11_2018}.  Three hot Jupiters
  are not shown because their radial velocity curves are better
  described as quadratic trends in time: HAT-P-17, WASP-8, and
  WASP-34.  Objects are ordered in the $y$ dimension by the absolute
  value of d$P$/d$t$.
	\label{fig:pdot_pop}
  \vspace{-0.3cm}
	}
\end{figure}
